---
title: "PRINCE: clinical and immune correlates with patternq"
subtitle: "prince-2022 on the Pattern Data Commons · R"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(patternq)
```

This notebook re-creates, on the Pattern Data Commons copy of the PRINCE trial,
the kinds of analyses in Padron et al., *Nat Med* 2022 (PRINCE: sotigalimab
and/or nivolumab with gemcitabine/nab-paclitaxel in first-line metastatic
pancreatic cancer): overall survival, circulating immune populations over
treatment, survival stratified at the median of a baseline biomarker, marker
expression on those populations, serum proteins, and tumor gene expression
signatures.

It is a demonstration of the `patternq` workflow, not a re-analysis.

> **Treatment arm is not in the `prince-2022` database.** The paper stratifies
> most analyses by arm (nivo/chemo, sotiga/chemo, sotiga/nivo/chemo). Arm is in
> the source data (`PICI0002_ph2_clinical.csv`, `Arm` / `Actual Arm`) but was
> not imported, so here every analysis pools patients across arms and groups
> by survival status at 1 year. P-values are unadjusted and exploratory.

The same notebook exists in Python (marimo), Clojure (Clay) and Julia in
`prince_demo.qmd` in [patternq](https://github.com/RCRF/patternq), `prince_demo.clj` in [patternq-clj](https://github.com/RCRF/patternq-clj) and
`prince_demo.qmd` in [PatternQ.jl](https://github.com/RCRF/PatternQ.jl).

## The dataset

Every dataset is its own database. `resolve_db` maps the stable dataset name
to the current database.

```{r}
db <- resolve_db("prince-2022")
set_db(db)
db
dataset_summary()
```

Which subjects were profiled by which assays:

```{r}
sa <- sample_assays()
plot_sample_overview(sa)
```

Samples by timepoint and specimen (blood draws at C1D1, C1D15, C2D1, C4D1;
tumor biopsies at screening):

```{r}
smp <- samples()
table(smp$timepoint_id, smp$sample_specimen)
```

## Outcomes

Best overall response, progression-free and overall survival (months) per
subject, and survival status at 1 year (the paper's primary endpoint is 1-year
OS). Subjects censored before 1 year have unknown status.

```{r}
outcomes <- subject_outcomes()
outcomes$status_1y <- survival_status(outcomes, "os", "os_event", at = 12)
status_levels <- c("alive at 1 year", "died within 1 year")
table(outcomes$status_1y, useNA = "ifany")
table(outcomes$bor, useNA = "ifany")
```

Overall survival (Fig. 2a pools arms here) and by best overall response:

```{r}
plot_survival(outcomes, "os", "os_event", title = "Overall survival", xlab = "months")
plot_survival(outcomes, "os", "os_event", group = "bor", title = "Overall survival by best overall response",
              xlab = "months")
```

Progression-free survival (Extended Data Fig. 2a):

```{r}
plot_survival(outcomes, "pfs", "pfs_event", title = "Progression-free survival", xlab = "months")
```

## Circulating immune populations

Frequencies of gated populations from the X50 (flow) and CyTOF panels, with
sample context (subject, timepoint) and 1-year survival status joined on.

```{r}
with_status <- function(tab) {
    tab <- add_sample_context(tab, include.subjects = FALSE)
    keep_provenance(merge(tab, outcomes[, c("subject_id", "status_1y")], by = "subject_id"), tab)
}
x50 <- with_status(measurements("percent-of-parent", "PICI X50 Immune Profiling"))
cytof <- with_status(rbind(
    measurements("percent-of-parent", "PICI CyTOF Immune Profiling"),
    measurements("percent-of-leukocytes", "PICI CyTOF Immune Profiling")))
blood_timepoints <- c("C1D1", "C1D15", "C2D1", "C4D1")
sort(unique(x50$cell_population))
```

Baseline (C1D1) value per subject for one population:

```{r}
baseline <- function(tab, population, timepoint = "C1D1") {
    b <- tab[tab$cell_population == population & tab$timepoint_id == timepoint, ]
    tapply(b$value, b$subject_id, mean)
}
```

### CD38+ effector memory CD8 T cells (Fig. 3a, c)

OS stratified by the pre-treatment frequency, above vs below the median:

```{r}
pop <- "CD38+ effector memory CD8+ T cells (% effector memory CD8 T cells)"
km <- survival_by_median(baseline(x50, pop), outcomes)
attr(km, "logrank")$p
plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
              title = "OS by baseline CD38+ EM CD8 T cells (median split)")
```

Frequencies before and on treatment, grouped by 1-year survival, with each
patient's trajectory:

```{r}
plot_by_timepoint(x50[x50$cell_population == pop, ], group = "status_1y", lines = TRUE,
                  timepoints = blood_timepoints, levels = status_levels,
                  title = "CD38+ EM CD8 T cells over treatment", ylab = "% of EM CD8 T cells")
```

### PD-1+CD39+ EM1 CD4 T cells (Fig. 3d) and PD-1+Tbet+ non-naive CD4 T cells (Fig. 4f, h)

```{r}
pops <- c("PD-1+CD39+ effector memory 1 CD4+ T cells (% of effector memory 1 CD4 T cells)",
          "PD-1+Tbet+ non-Naive CD4+ T cells (% of CD4 not naive T cells)")
htmltools::tagList(lapply(pops, function(pop) {
    km <- survival_by_median(baseline(x50, pop), outcomes)
    plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
                  title = paste("OS by baseline", sub(" \\(.*", "", pop)))
}))
```

### Cross-presenting dendritic cells (Fig. 4b)

```{r}
pop <- "CD1C+ CD141+ DC (% of leukocytes)"
km <- survival_by_median(baseline(cytof, pop), outcomes)
plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
              title = "OS by baseline CD1c+ CD141+ DCs (CyTOF)")
```

### Proliferating T cells over treatment (Extended Data Fig. 3a)

Change relative to C1D1 (log2 fold change of the frequency):

```{r}
ki67 <- change_from_baseline(x50[x50$cell_population == "Ki67+ Non-Naive CD8 T Cells", ],
                             baseline = "C1D1", method = "log2_ratio", pseudocount = 0.001)
plot_by_timepoint(ki67, value = "change", group = "status_1y", lines = TRUE,
                  timepoints = blood_timepoints, levels = status_levels,
                  title = "Ki-67+ non-naive CD8 T cells, change from C1D1",
                  ylab = "log2 fold change vs C1D1")
```

### Screening all baseline populations

Every X50 and CyTOF population at C1D1, median split, log-rank p (the paper's
Fig. 5 summarizes such associations per arm):

```{r}
screen <- function(tab, panel) {
    do.call(rbind, lapply(unique(tab$cell_population), function(pop) {
        km <- survival_by_median(baseline(tab, pop), outcomes)
        lr <- attr(km, "logrank")
        if (is.na(lr$p)) return(NULL)  # no baseline values for this population
        hr <- lr$observed / lr$expected
        data.frame(panel = panel, cell_population = pop, n = nrow(km), p = lr$p,
                   higher_is = if (isTRUE(hr[["high"]] < hr[["low"]])) "longer survival" else "shorter survival")
    }))
}
assoc <- rbind(screen(x50, "X50"), screen(cytof, "CyTOF"))
assoc <- assoc[order(assoc$p), ]
knitr::kable(head(assoc, 12), digits = 3, row.names = FALSE)
```

## Marker expression on populations (Extended Data Figs. 7b, 8b)

Median channel values of markers on CD38+ non-naive CD8 T cells, pre-treatment,
per patient, z-scored per marker, patients annotated by 1-year survival:

```{r}
mcv <- with_status(measurements("median-channel-value", "PICI X50 Immune Profiling"))
cd8 <- mcv[mcv$cell_population == "CD38+ non-naive CD8+ T cells (% non-naive CD8 T cells)" &
           mcv$timepoint_id == "C1D1", ]
m <- to_matrix(cd8, row = "epitope_id", col = "subject_id")
status <- setNames(outcomes$status_1y, outcomes$subject_id)
plot_heatmap(m, scale = "row", col.groups = status,
             title = "Markers on CD38+ non-naive CD8 T cells (C1D1, z-scored MFI)")
```

CCR7+ CD11b+ CD27- B cells (CyTOF) on treatment (C1D15):

```{r}
mcv_b <- with_status(measurements("median-channel-value", "PICI CyTOF Immune Profiling"))
b <- mcv_b[mcv_b$timepoint_id == "C1D15", ]
plot_heatmap(to_matrix(b, row = "epitope_id", col = "subject_id"), scale = "row", col.groups = status,
             title = "Markers on CCR7+ CD11b+ CD27- B cells (C1D15, z-scored MFI)")
```

## Serum proteins (Extended Data Fig. 3c–f)

Olink NPX is on a log2 scale, so change from C1D1 is a difference:

```{r}
olink <- with_status(measurements("olink-npx", "PICI Olink Proteomics"))
olink_change <- change_from_baseline(olink, baseline = "C1D1", method = "difference")
proteins <- c(IFNG = "IFNG", `PD-1` = "PDCD1", CXCL9 = "CXCL9", CXCL10 = "CXL10")  # label = Olink id
htmltools::tagList(Map(function(label, id)
    plot_by_timepoint(olink_change[olink_change$epitope_id == id, ], value = "change", group = "status_1y",
                      timepoints = blood_timepoints, levels = status_levels,
                      title = paste(label, "change from C1D1"), ylab = "log2 NPX change"),
    names(proteins), proteins))
```

## Tumor: gene expression signatures and multiplex IF

Pre-treatment tumor RNA-seq signature scores, clustered, patients annotated by
1-year survival (Extended Data Figs. 4a, 6a). The RNA-seq data comes from the
cBioPortal / iAtlas release, which labels the pre-treatment timepoint `0`
(the PICI modalities use `screening` / `C1D1`):

```{r}
sig <- with_status(measurements("nanostring-signature-score", "cBioPortal Gene Expression Signatures"))
sig_pre <- sig[sig$timepoint_id %in% c("screening", "0"), ]
sm <- to_matrix(sig_pre, row = "signature", col = "subject_id")
plot_heatmap(sm, scale = "row", col.groups = status,
             title = "Tumor gene expression signatures (pre-treatment, z-scored)")
```

OS by the Th1 signature (Extended Data Fig. 6b):

```{r}
th1 <- sig_pre[sig_pre$signature == "Bindea_Th1_Cells", ]
km <- survival_by_median(tapply(th1$value, th1$subject_id, mean), outcomes)
plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
              title = "OS by tumor Th1 signature (median split)")
```

OS by intratumoral iNOS+ macrophages from multiplex IF (Extended Data Fig. 4c):

```{r}
vectra <- with_status(measurements("percent-of-total-cells", "PICI Vectra Multiplex IF"))
inos <- vectra[vectra$cell_population == "iNOS+ macrophages (% of total cells)" &
               vectra$timepoint_id == "screening", ]
km <- survival_by_median(tapply(inos$value, inos$subject_id, mean), outcomes)
plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
              title = "OS by tumor iNOS+ macrophages (median split)")
```

## Tumor mutations

```{r}
v <- variants()
plot_mutation_landscape(v, n.genes = 20, title = "PRINCE mutation landscape (top 20 genes)")
```

## Provenance

Every result records the database and the database basis t it was read at:

```{r}
str(provenance(x50))
```
