---
title: "patternq tutorial"
subtitle: "Querying and analyzing the Pattern Data Commons · R"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

The basic patternq workflow, step by step. The same steps, with the same
datasets and outputs, are in `tutorial.qmd` in [patternq](https://github.com/RCRF/patternq) (marimo),
`tutorial.clj` in [patternq-clj](https://github.com/RCRF/patternq-clj) (Clay) and `tutorial.qmd` in [PatternQ.jl](https://github.com/RCRF/PatternQ.jl) (Julia).

| step | what |
|---|---|
| 1 | connect and find a dataset |
| 2 | explore a dataset: assays, measurement sets, samples, subjects |
| 3 | write a Datalog query |
| 4 | canned queries: variants and gene expression |
| 5 | any measurement, reshaped and joined with sample context |
| 6 | clinical outcomes and survival |
| 7 | a sample against a reference cohort |
| 8 | expression change between two samples |

## 1. Connect and find a dataset

Set `PATTERNQ_API_KEY` (and optionally `PATTERNQ_ENDPOINT`) in the environment,
or call `set_token()` / `set_query_server()`.

```{r}
library(patternq)
query_server()
datasets <- list_datasets()
head(datasets[, c("dataset", "db", "patient_count", "sample_count")])
```

Every dataset is its own database. Dataset names are stable; database names
change when a dataset is re-imported, so resolve them:

```{r}
h37001 <- resolve_db("H37001")
uvm <- resolve_db("tcga-uvm")
c(h37001, uvm)
set_db(h37001)
```

## 2. Explore a dataset

```{r}
dataset_summary()
measurement_types("RSEM RNASeq")
samples()
subjects()
```

## 3. Write a Datalog query

Queries are Datomic Datalog in the query service's JSON form: variables are
`"?x"`, attributes `":ns/name"`, each clause a vector.

```{r}
q <- dq(find = c("?sample-id", "?site"),
        where = list(c("?s", ":sample/id", "?sample-id"),
                     c("?s", ":sample/freetext-anatomic-site", "?site")))
do_query(q)
```

Inputs bind with `in` / `args`; a vector binds a collection:

```{r}
q <- dq(find = c("?sample-id", "?hgnc", "?vaf"),
        `in` = list(c("?hgnc", "...")), args = list(c("BAP1", "GNAQ", "SF3B1")),
        where = list(c("?g", ":gene/hgnc-symbol", "?hgnc"),
                     c("?v", ":variant/gene", "?g"),
                     c("?m", ":measurement/variant", "?v"),
                     c("?m", ":measurement/vaf", "?vaf"),
                     c("?m", ":measurement/sample", "?s"),
                     c("?s", ":sample/id", "?sample-id")))
do_query(q)
```

Every result records where it came from:

```{r}
str(provenance(do_query(q)))
```

## 4. Variants and gene expression

```{r}
v <- variants()
nrow(v)
head(v[v$impact == "high", c("sample_id", "hgnc_symbol", "HGVSp", "vaf")])
plot_vaf_histogram(v, samples = c("H37001-001", "H37001-006"))
```

```{r}
gx <- gene_expression(genes = c("BAP1", "GNAQ", "PRAME", "PMEL", "MLANA"), measurement = "tpm")
head(gx)
plot_gene_expression(gx, title = "Melanocytic genes (TPM)", ylab = "TPM", log = TRUE)
```

Copy number queries always take a subset (genes, samples or subjects):

```{r}
cnv_segments(genes = c("BAP1", "GNAQ"))
```

## 5. Any measurement, reshaped and joined with context

`measurements()` works for any measurement attribute and detects what each
measurement is of (gene, cell population, epitope, ...). Here: CyTOF cell
population frequencies in PRINCE.

```{r}
prince <- resolve_db("prince-2022")
measurement_types("PICI CyTOF Immune Profiling", db = prince)
cy <- measurements("percent-of-parent", "PICI CyTOF Immune Profiling", db = prince)
head(cy)
```

Wide (samples x populations):

```{r}
m <- to_matrix(cy, col = "cell_population")
dim(m)
```

With sample and subject context:

```{r}
cy <- add_sample_context(cy, db = prince)
head(cy[, c("sample_id", "subject_id", "timepoint_id", "cell_population", "value")])
plot_by_timepoint(cy[cy$cell_population == "HLA-DR+ Non-Naive CD8 T Cells", ],
                  timepoints = c("C1D1", "C1D15", "C2D1", "C4D1"), lines = TRUE,
                  title = "HLA-DR+ non-naive CD8 T cells", ylab = "% of parent")
```

## 6. Clinical outcomes and survival

```{r}
outcomes <- subject_outcomes(prince)
head(outcomes)
plot_survival(outcomes, "os", "os_event", group = "bor", xlab = "months",
              title = "Overall survival by best overall response")
```

Survival split at the median of a baseline biomarker, with a log-rank test:

```{r}
base <- cy[cy$cell_population == "HLA-DR+ Non-Naive CD8 T Cells" & cy$timepoint_id == "C1D1", ]
km <- survival_by_median(tapply(base$value, base$subject_id, mean), outcomes)
attr(km, "logrank")$p
plot_survival(km, group = "group", levels = c("low", "high"), xlab = "months",
              title = "OS by baseline HLA-DR+ non-naive CD8 T cells")
```

## 7. A sample against a reference cohort

Place a patient sample within a reference cohort's distribution, gene by gene
(z-score and percentile on log2(1 + TPM)). Only compare comparable
measurements, and check `cohort_observed`.

```{r}
genes <- c("BAP1", "GNAQ", "PRAME", "MITF", "PMEL", "TYR", "MLANA", "CD8A", "GZMB")
cmp <- compare_to_cohort("H37001-003", db = h37001, cohort.db = uvm, genes = genes)
cmp[, c("hgnc_symbol", "value", "z", "percentile", "cohort_observed")]
examine_geneset(genes, samples = c("H37001-003", "H37001-001"), db = h37001,
                cohort.dbs = c("TCGA-UVM" = uvm), title = "H37001 vs TCGA-UVM")
```

## 8. Expression change between two samples

```{r}
change <- compare_samples("H37001-003", "H37001-001", db = h37001)
head(change)
plot_ma(change)
```
